---
title: "Mathematical description of SFGP models"
author: "Juho Timonen"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Mathematical description of SFGP models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(sfgp)
```

# Longitudinal model

This section describes mathematically the model corresponding to 
the `TSModel` class.

## Observation model
The likelihood of observation $i$ is
$$
\log(y_i + \delta) \sim \mathcal{N}(f_i, \sigma^2),
$$
where 

$$
f_i = f(\mathbf{x}_i)= \sum_{j=1}^J f^{(j)}(\mathbf{x}_i)
$$
is the expected log tumor size, $\sigma$ is an unknown parameter, and 
the $\delta$ value is a constant. We use $y_i$ to denote the observed
longitudinal measurement, and $\mathbf{x}_i$ the corresponding covariate
vector. Typically, $\mathbf{x}_i$ includes a subject
identifier $\text{id}_i$, measurement time $t_i$, treatment arm $\text{arm}_i$ 
and possible other factors or continuous covariates.
The functions $f^{(j)}$, $j=1, \ldots, J$ are the additive function
components and they are detailed next.

## Model terms


```{r}
create_term <- function(term_str) {
  m <- TSModel$new(y ~ ., baseline = term_str, compile = FALSE)
  m$term_list$terms[[1]]
}
```

The model can have a different number $J$ of terms, depending on the 
`formula` that is given.

### `GroupedOffsetTerm`

```{r}
term <- create_term("offset(g)")
print(term)
cat(term$latex())
```

The `GroupedOffsetTerm` for a grouping variable $g$ is
$$
f^{\text{BAS}} \left(\text{g} \mid \mathbf{c}_{0}\right)
$$
where $\mathbf{c}_{0}$ is an unknown parameter vector containing one
parameter for each level of the grouping variable. These parameters have
independent priors.

### `HierOffsetTerm`

```{r}
term <- create_term("offset(g | h)")
print(term)
cat(term$latex())
```

The `HierOffsetTerm` for grouping variables $z$ and $h$ is
$$
f^{\text{BAS}} \left(\text{g}, \text{h} \mid \mathbf{c}_{0}\right)
$$
where $\mathbf{c}_{0}$ is again an unknown parameter vector containing one
parameter for each level of the grouping variable. The difference to
`GroupedOffsetTerm` is that the prior is hierarchical. It is assumed that
for the value of $h = h[g]$ is always determined by $g$. So for example $g$ 
can be the subject identifier and $h[g]$ the treatment arm for that subject. The 
prior is parametrized so that for group $g$ the parameter is

$$
\mathbf{c}_{0, g} = z_{g} \cdot \sigma_{h[g]} + \mu_{h[g]}
$$
where by default 
$\mu_{h} \sim \mathcal{N}(0,1), \sigma_{h} \sim \mathcal{N}(0,5)$ independently
over all $h$ and $z_g \sim \mathcal{N}(0,1)$ independently over all $g$.

```{r}
cat(term$stancode_model())
```

### Shared GP terms (`GPTerm`)

```{r}
term <- create_term("gp(x)")
print(term)
cat(term$latex())
```

Let $x$ be some continuous variable. We use
$$
f^{\text{GP}}(x \mid \alpha, \ell) \sim \mathcal{GP}(0,\ k_{\text{EQ}}(x,x' \mid \alpha, \ell)).
$$
to denote that a function is modeled as a zero-mean Gaussian process (GP),
with the exponentiated quadratic kernel function
$$
k_{\text{EQ}}(x, x' \mid \alpha, \ell) = \alpha^2 \exp\left( -\frac{(x-x')^2}{2\ell^2}\right),
$$
where the parameters $\alpha > 0$ and $\ell > 0$ are the kernel magnitude and 
lengthscale parameters, respectively. For GP terms, we use a Hilbert space
basis function approximation
$$
f^{\text{GP}}(x \mid \alpha, \ell) \approx f^{\text{HSGP}}(x \mid \mathbf{\xi}, \alpha, \ell, B, L) = \sum_{b=1}^{B} \xi_b S_{\text{EQ}}\left(\sqrt{\lambda_{L,b}} \mid \alpha, \ell \right) \phi_{L,b}(x),
$$
 where the auxiliary parameters 
$\mathbf{\xi} = \left\{\xi_1, \ldots, \xi_B\right\}$ are a priori standard
normal distributed $\xi_b \sim \mathcal{N}(0,1)$. This approximation approaches
the exact GP in the limit $L \rightarrow \infty$, $B \rightarrow \infty$. The 
basis functions 
$$
\phi_{L,b}(x) = \frac{1}{\sqrt{L}} \sin \left( \frac{\pi b (x + L)}{2L}\right)
$$
are eigenfunctions of a certain Dirichlet boundary value problem,
and $\lambda_{L, b} = \left(\frac{\pi b}{2 L} \right)^2$ are their
corresponding eigenvalues. The function
$$
S_{\text{EQ}}(\omega \mid \alpha, \ell) = \alpha^2 \ell \sqrt{2 \pi} \exp \left(-\frac{\ell^2 \omega^2 }{2}  \right)
$$
is the spectral density of the exponentiated quadratic kernel.The approximation 
domain size $L$ and number of basis functions $B$ are 
constants that are set beforehand. The approximation is accurate only on
the interval $x \in (-L, L)$.

The unknown parameters of a shared GP term are $\alpha, \ell$ and 
$\mathbf{\xi} \in \mathbb{R}^B$.

### Group-specific GP terms (`GPTerm`)

```{r}
term <- create_term("gp(x, z)")
print(term)
cat(term$latex())
```

Let $x$ be some continuous variable and $z$ some grouping factor, which can
without loss of generality take integer values $z \in \{1, \ldots, G\}$. A 
group-specific GP term is
$$
f^{\text{HSGP}}(x, z \mid \mathbf{\xi}_1, \ldots, \mathbf{\xi}_G, \alpha, \ell, B, L) = 
\begin{cases}
f^{\text{HSGP}}(x \mid \mathbf{\xi}_1, \alpha, \ell, B, L) & \text{if } z = 1\\
\vdots \\
f^{\text{HSGP}}(x \mid \mathbf{\xi}_G, \alpha, \ell, B, L) & \text{if } z = G
\end{cases}
$$
which approximates the GP
$$
f^{\text{GP}}(x,z) \sim \mathcal{GP}(0,\ k_{\text{CAT}}(z,z') \cdot k_{\text{EQ}}(x,x')),
$$
where $k_{\text{CAT}}(z,z') = 1$ if $z=z'$ and zero otherwise. The unknown
parameters of a group-specific GP term are $\alpha, \ell$ and a different vector 
$\mathbf{\xi}_g \in \mathbb{R}^B$ for each group $g=1, \ldots, G$.

### The Stein-Fojo term (`SFTerm`)

```{r}
term <- create_term("sf(x)")
print(term)
cat(term$latex())
```


The Stein-Fojo (SF) function is
$$
f^{\text{SF}}(x \mid k_g, k_s) = \exp(k_g x) + \exp(-k_s x) - 1,
$$
where $k_g, k_s$ are unknown growth and shrinkage parameters, respectively. 

### `FormulaSFTerm`

In the `FormulaSFTerm`, the parameters $k_g$ and $k_s$ can depend on other
covariates using all the other terms defined earlier.

```{r}
term <- create_term("sff(t | kg ~ offset(id_kg) + gp(x_kg), ks ~ offset(id_ks))")
print(term)
cat(term$latex())
cat(term$term_list_kg$latex())
cat(term$term_list_ks$latex())
```

The formulae defined for $k_g$ and $k_s$ mean that
$$
k_{g}(\mathbf{x})) = \exp(\log(C_g) + f_{g}(\mathbf{x}))
$$
and

$$
k_{s}(\mathbf{x})) = \exp(\log(C_s) + f_{s}(\mathbf{x}))
$$
where $C_g, C_s$ are unknown parameters and $f_g$, $f_s$ are functions
defined by the formulae.


Short-hand for the basic hierarchical SF term is

```{r}
term <- create_term("sf(t, id | arm)")
print(term)
cat(term$latex())
cat(term$term_list_kg$latex())
cat(term$term_list_ks$latex())
```

## Examples

### The SF-only model

In the basic SF-only model with the formula we have
$\mathbf{x} = \{t, \text{id}, \text{arm}\}$, $J = 2$ and

```{r}
mod1 <- TSModel$new(y ~ sf(t, id | arm), compile = FALSE)
cat(mod1$term_list$latex())
```

\begin{align}
f^{(1)}(\mathbf{x}) &= f^{\text{log-SF}} \left(\text{t} \mid \mathbf{k}_{g}, \mathbf{k}_{s}\right)\\ 
  f^{(2)}(\mathbf{x}) &= f^{\text{BAS}} \left(\text{id} \mid \mathbf{c}_{0}\right)
\end{align}.

Notice that in this case

$$
\exp(f(\mathbf{x})) = \exp\left(\log \text{c}_{0,\text{id}} + \log f^{\text{SF}}(t \mid k_{g, \text{id}}, \mid k_{s, \text{id}}) \right) = \text{c}_{0,\text{id}} \left( \exp(k_{g, \text{id}} \cdot t) + \exp(-k_{s, \text{id}} \cdot t) - 1 \right)
$$
which is the original formulation of the SF tumor size model.

### An SF+GP model

An example of an SF+GP model with $\mathbf{x} = \{t, \text{id}, \text{arm}\}$,
$J = 4$ is 

```{r}
mod2 <- TSModel$new(y ~ sf(age, id | arm) + gp(age) + gp(age, arm), compile = FALSE)
cat(mod2$term_list$latex())
```

It has the terms
 
\begin{align}
f^{(1)}(\mathbf{x}) &= f^{\text{log-SF}} \left(\text{age} \mid \mathbf{k}_{g}, \mathbf{k}_{s}\right)\\ 
  f^{(2)}(\mathbf{x}) &= f^{\text{HSGP}} \left(\text{age} \mid \mathbf{\xi}_{\text{age}}, \alpha_{\text{age}}, \ell_{\text{age}}, B_{\text{age}}, L_{\text{age}}\right)\\ 
  f^{(3)}(\mathbf{x}) &= f^{\text{HSGP}} \left(\text{age}, \text{arm} \mid \mathbf{\xi}^{(1)}_{\text{age} \times \text{arm}}, \ldots, \mathbf{\xi}^{(G_{arm})}_{\text{age} \times \text{arm}}, \alpha_{\text{age} \times \text{arm}}, \ell_{\text{age} \times \text{arm}}, B_{\text{age} \times \text{arm}}, L_{\text{age} \times \text{arm}}\right)\\ 
  f^{(4)}(\mathbf{x}) &= f^{\text{BAS}} \left(\text{id} \mid \mathbf{c}_{0}\right)
\end{align}

where $G_{\text{arm}}$ is the number of treatment arms.
