---
title: "survival-model-check"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
author: Jacqueline Buros Novik and Juho Timonen
vignette: >
  %\VignetteIndexEntry{survival-model-check}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setup

```{r setup}
# library(devtools)
# Sys.setenv(GITHUB_PAT = "")
# devtools::install_github("sambrilleman/simjm")
library(simjm)
library(sfgp)
library(ggplot2)
library(survminer)
library(survival)
library(tidyverse)
```

# Model

```{r}
tsm <- TSModel$new(y ~ offset(z) + gp(t), compile = F)
m <- JointModel$new(lon = tsm)
print(m)

format_lon_data <- function(data) {
  dat_lon <- data$Long1[, c("id", "Z1", "Z2", "tij", "Yij_1")]
  dat_lon$id <- as.factor(dat_lon$id)
  dat_lon$Z1 <- as.factor(dat_lon$Z1)
  colnames(dat_lon) <- c("id", "z", "x", "t", "y")
  dat_lon
}
format_tte_data <- function(data) {
  dat_tte <- data$Event[, c("id", "Z1", "Z2", "eventtime", "status")]
  dat_tte$id <- as.factor(dat_tte$id)
  dat_tte$Z1 <- as.factor(dat_tte$Z1)
  dat_tte$status <- as.logical(dat_tte$status)
  colnames(dat_tte) <- c("id", "z", "x", "t", "event")
  dat_tte
}
format_data <- function(data, num_points = 30) {
  dat_lon <- format_lon_data(data)
  # truncating at values less than 0.01
  dat_lon$y <- if_else(dat_lon$y < 0.01, 0.01, dat_lon$y)
  dat_tte <- format_tte_data(data)
  list(
    lon = dat_lon,
    tte = dat_tte,
    grid = create_jm_grid(dat_tte, dat_lon,
      num_points = num_points,
      id_var = "id"
    )
  )
}
plot_h0 <- function(fit) {
  h0 <- fit$h0()
  # We do this to avoid grouping by id in the plot
  inp <- h0$get_input()
  inp <- inp[, "t", drop = FALSE]
  h0_new <- FunctionDraws$new(inp, h0$get_output(), h0$get_name())

  h0_new$plot()
}
plot_params <- function(post_params, true_params) {
  ggplot(post_params, aes(x = .value, group = .variable)) +
    tidybayes::stat_halfeye() +
    facet_wrap(~.variable, scale = "free") +
    geom_vline(aes(xintercept = .value), data = true_params, color = "firebrick")
}
plot_corr <- function(post_params) {
  ggplot(
    post_params |> tidyr::spread(.variable, .value),
    aes(x = assoc, y = h0_lambda)
  ) +
    geom_point() +
    ggtitle("Correlation between posterior draws for assoc & h0_lambda")
}
```

# Scenario 1: No association, Intercept-only, Weibull hazard

First we simulate data with no covariate effects on the hazard & no association.

This is a Weibull model with an intercept only and gamma set to 1, so it's equivalent to an exponential model with constant hazard. 

The question is whether we can recover the shape parameter.

```{r sim01, fig.width=7, fig.height=5}
params <- list(h0_gamma = 1, h0_lambda = 0.2)
data <- simjm(
  assoc = "null",
  n = 30, basehaz = "weibull",
  fixed_trajectory = "linear",
  random_trajectory = "none",
  b_sd = 1,
  betaEvent_intercept = c(log(params$h0_lambda)),
  betaEvent_assoc = c(0),
  betaEvent_binary = c(0),
  betaEvent_aux = c(params$h0_gamma),
  max_fuptime = 50,
  betaEvent_continuous = c(0),
  seed = 10
)
params1 <- attr(data, "params")
plot(data)

# plot simulated event data
sfit <- survfit(Surv(eventtime, status) ~ 1, data = data$Event)
ggsurvplot(sfit, data = data$Event)
```


## Viewing the data

Next, let's plot the raw longitudinal data

```{r, fig.width=7, fig.height=5}
data$Long1 |>
  ggplot(aes(x = tij, y = Yij_1, group = factor(id), colour = factor(id))) +
  geom_line()
```



```{r}
dat <- format_data(data)
case1_fit <- fit <- m$fit(
  data_lon = dat$lon, data_tte = dat$tte, data_grid = dat$grid,
  chains = 1
)
```

## Summarize results


### Baseline and instant hazard

Note that the "assoc" term (and other betas) can absorb some of the "lambda" 
term, particularly if predictors are not centered.


```{r, fig.width=7, fig.height=5}
plot_h0(fit)
fit$plot_hazard()
fit$plot(facet_by = "id")
```

### Association and parameters 

```{r, fig.width=7.2, fig.height=3}
print(fit$draws("assoc")) # association
true_params <- tibble::tibble(!!!params, assoc = 0) |>
  tidyr::gather(".variable", ".value")
stan_fit <- fit$get_stan_fit()
post_params <- tidybayes::gather_draws(stan_fit, h0_lambda, h0_gamma, assoc)
plot_params(post_params, true_params)
```

Note that the "assoc" term (and other betas) can absorb some of the "lambda" 
term, particularly if predictors are not centered.

```{r, fig.width=5.2, fig.height=4}
plot_corr(post_params)
```

### Survival function

```{r, fig.width=7, fig.height=5}
p <- fit$predict(data_tte = dat$grid)
p$survival_function()$plot()
```


# Scenario 2: With association 

```{r sim02, fig.width=7, fig.height=5}
params <- list(h0_lambda = 0.01, h0_gamma = 1)
true_assoc <- 0.35
data <- simjm(
  n = 30, basehaz = "weibull",
  fixed_trajectory = "linear",
  random_trajectory = "none",
  b_sd = 1,
  betaEvent_intercept = c(log(params$h0_lambda)),
  betaEvent_assoc = c(true_assoc),
  betaEvent_binary = c(0),
  betaEvent_continuous = c(0),
  betaEvent_aux = c(params$h0_gamma),
  max_fuptime = 50,
  max_yobs = 20,
  seed = 100
)
print(attr(data, "params"))
plot(data)
# plot_traj(yvar = "Yij_1", data = data$Long1)
```

## Viewing the data

Next, let's plot the raw longitudinal data

```{r, fig.width=7, fig.height=5}
data$Long1 |>
  ggplot(aes(x = tij, y = Yij_1, group = factor(id), colour = factor(id))) +
  geom_line()
```



```{r}
dat <- format_data(data)
case2_fit <- fit <- m$fit(
  data_lon = dat$lon, data_tte = dat$tte, data_grid = dat$grid,
  chains = 1
)
```

## Summarize results


### Baseline and instant hazard


```{r, fig.width=7, fig.height=5}
plot_h0(fit)
fit$plot_hazard() + scale_y_log10()
fit$plot(facet_by = "id")
```

### Association and parameters 

```{r, fig.width=7.2, fig.height=3}
print(fit$draws("assoc")) # association
true_params <- tibble::tibble(!!!params, assoc = true_assoc) |>
  tidyr::gather(".variable", ".value")
stan_fit <- fit$get_stan_fit()
post_params <- tidybayes::gather_draws(stan_fit, h0_lambda, h0_gamma, assoc)
plot_params(post_params, true_params)
```

```{r, fig.width=5.2, fig.height=4}
plot_corr(post_params)
```
### Survival function

```{r, fig.width=7, fig.height=5}
p <- fit$predict(data_tte = dat$grid)
p$survival_function()$plot()
```



# Used Stan code
```{r}
cat(m$create_stancode())
```
